---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# wmwm

<!-- badges: start -->
<!-- badges: end -->

This package wmwm is to perform the two-sample Wilcoxon-Mann-Whitney test in the presence of missing data. The Wilcoxon-Mann-Whitney test, also known as Wilcoxon rank sum test or Mann-Whitney U test, is a non-parametric two-sample hypothesis testing method. Suppose `X` and `Y` are sets of univariate, distinct, real-valued samples assumed randomly selected from two populations. The null hypothesis of the test is that
the two populations are the same. If both `X` and `Y` are fully observed, stats::wilcox.test() in R is for the Wilcoxon-Mann-Whitney test.

While only subsets of `X` and `Y` are observed, denote observed samples in `X` and `Y` as `X'` and `Y'`, respectively. The Wilcoxon-Mann-Whitney test can not be carried out directly. The default method in stats::wilcox.test() is to ignore all unobserved samples and perform the test between `X'` and `Y'`.
Unfortunately, this practice of ignoring unobserved samples is rarely valid and can potentially leads to a higher type I error than a pre-specified significance level.

This package provides a method for conducting two-sample Wilcoxon-Mann-Whitney test in the presence of missing data by computing the bounds of the Wilcoxon-Mann-Whitney test statistic and p-values.
Then, the null hypothesis should be rejected if all test statistics are significant, alternatively,
if all p-values are smaller than a pre-specified significance level. This method controls the type I error under a pre-specified significance level regardless of the values of missing data. This package also allows the test if `X` and `Y` are multisets consisting of real-valued, potentially tied samples.

## Installation

You can install the development version of wmwm from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Yijin-Zeng/Wilcoxon-Mann-Whitney-Test-with-Missing-data")
```

## Example

As a very simple example, the following shows how we can perform the Wilcoxon-Mann-Whitney test in the presence of missing data by computing the bounds of the statistic and the p-value. To start, let us randomly generate our samples.

```{r}
   library(wmwm)
   set.seed(1) 
   ## tested samples
   X <- stats::rnorm(100,0,1)
   X[1:5] <- NA      
   Y <- stats::rnorm(60,1,1)
   Y[1:5] <- NA
   X[1:6]
   Y[1:6]
```

In `X` and `Y`, each unobserved sample is shown as NA. Suppose a two-sided Wilcoxon-Mann-Whitney test between `X` and `Y` with a significance level $\alpha = 0.05$ is of interest. The testing can be done by runing: 

```{r example}
   wmwm.test(X,Y)
```

Subsequently, since `bounds.pvalue` shows the p-value is bounded lower than 0.0149, which is smaller than the pre-specified significance level $\alpha = 0.05$, the null hypothesis should then be rejected.

Additionally, the `bounds.statistic` gives the bounds of the Wilcoxon test statistic; `alternative` specifies the alternative hypothesis; `description.bounds` describes the assumptions underlying the bounds.  

## Optional choices

This function provides other options of the bounds. For example, instead of bounding the p-value using normal approximation, we may want to bound the exact p-value. This can be done by running:
```{r}
   wmwm.test(X,Y, exact = TRUE)
```

In some cases, we need to deal with tied samples, i.e. two or more observations may be equal. This function deal with ties using mid-rank practice automatically if observed samples containing ties observations.
```{r}
   X <- stats::rpois(100,1)
   X[1:5] <- NA      
   Y <- stats::rpois(60,2)
   Y[1:5] <- NA
   X[1:6]
   Y[1:6]
   wmwm.test(X,Y)
```
If we know that the unobserved samples can not be smaller than the minimum of observed samples, (in this case, can not smaller than 0),  we can specify `lower.boundary = 'equal'` to make the bounds potentially tighter.
```{r}
   min(c(X[!is.na(X)], Y[!is.na(Y)]))
   wmwm.test(X,Y,lower.boundary = 'equal')
```


## References 

Mann, Henry B., and Donald R. Whitney. "On a test of whether one of two random variables is stochastically larger than the other." The annals of mathematical statistics (1947): 50-60.
 
Lehmann, Erich Leo, and Howard J. D'Abrera. Nonparametrics: statistical methods based on ranks. Holden-day, 1975.
  
Schafer, Joseph L., and John W. Graham. "Missing data: our view of the state of the art." Psychological methods 7.2 (2002): 147.
